Suitability prevalence area index in late quaternary explains genetic diversity in Tassel eared Squirrels

### 2023-07-29 manuscript_english.tex
changes in manuscript

@@

Our results reveal that suitability prevalence corresponds to the
fixation index of \emph{S. aberti} populations with respect to a source
population. Populations closer to the historical endemic area present a
higher genetic diversity and a lower \(F_{st}\) value. Finally, this
study allows us to add a biogeographic explanation to the results
obtained with population genetic methods and to generate maps of this
structure as tools to support conservation with a perspective that
integrates both population genetics and historical patterns of species
distribution.

@@

For a given region \(\mathbf{G}\) in te geography, at any point in time,
we have a set of existing environmental conditions
\(\mathbf{E}(\mathbf{G}, t) = \Big \{ \vec{e}_i; \mathbf{G}, t\Big \}\).
We define \(\vec{e}_i =(e_1, e_2, ..., e_v )_i\) as a vector of
environmental variables, that is, a point in the space of \(v\)
environmental variables \(\mathbf{E}(\mathbf{G}, t)\). Then
\(R_0(\vec{e}_i)\) is a function that relates fitness to environment,
that is, for a species with non-overlapping generations, the
population's net reproductive rate \(R_0\) is a measure of fitness that
depends on environmental conditions. If we take the set of places in
environmental space where these values are greater than a threshold, we
can define a volume in multidimensional space such that \(R(\vec{e})\)
is always positive and greater than this threshold, that is
\(\mathbf{N}_F=\{\vec{e} | R_0(\vec{e}) \ge k \}\). i. e. the set of
environmental conditions for which values of fitness are capable of
sustaining populations. We call
\(\mathbf{N}_F \subseteq \mathbf{E}(\mathbf{G}, t)\). This is a way of
expressing the fundamental niche as a function of environmental
conditions. So we can correspong each point of geography to
environmental space, with different environmental conditions and
different growth rates. However all sites within the region defined by
\(\mathbf{N}_F\) will have viable populations and positive rates
depending on time because we assume that the environment is dynamic and
each geographic point is static, so for each site in geographic space
this growth rate depends on the environment and therefore is different
at each time: \(R_0 = R_0(\vec{e}(t)) \}\). Thus locally the growth rate
will be related to the local population size as long as it is within the
boundary defined by \(\mathbf{N}_F\).

@@

This set can be seen as a convex volume in environmental space and we
relate this growth rate to an environmental suitability function bounded
between \(0\) and \(1\), for example, for a unimodal multivariate pdf,
it would be equivalent to the distance from the maximum of the function.
That is, the fitness-related environmental suitability can be any
function such that it is \(0\) if \(R_0(\vec{e}_i) > 1\) and \(1\) when
\(R_0\) is evaluated at the expected value of \(\vec {e}(t)\) for all
values of the set \(\mathbf{N}_F\) where \(R_0(\vec{e}_i) > 1\). In a
practical way, instead of \(\mathbf{N}_F\), we approximated
\(\mathbf{N}_F^{*}(\mathbf{G, t}) = \mathbf{E} \cap \mathbf{N}_F\) (the
existing niche) from known presence points in the geography
\(\mathbf{G}\) obtained from emprical observations stored in database,
and map to \(\mathbf{E}\) in order to numerically fitting a convex
volume of the environmental space that enclose the known information.

The objective of our approach is not to know the function of \(R_0\) but
rather to relate this local variation and the function \(\mathbf{N}_F\)
with a function \(S_i(\vec{e}_i)\) of environmental suitability that can
be estimated from known methods of applied mathematics and machine
learning (for example the case of Maxent, Phillips, Anderson, and
Schapire (2006)).

Population size locally is a function of environmental suitability,
where if suitability is \(0\) the population is \(0\) and if suitability
is \(1\) the population has a maximum growth rate \(R_i(\vec{e}_i)\) \[
N_i \propto  S_i(\vec{e}_i)
\]

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,height=\textheight]{all_figures/figure_1.png}
\caption{Hypothetical niche in environmental space. An arbitrary point
moves within the environmental space entering and exiting the volume
(niche) where the intrinsic growth rate is positive.}
\end{figure}

Once we have an approximation of this volume, we consider that each
point in the geography is static and has a trajectory in the
environmental space and for each trajectory the fitness function changes
and therefore the suitability function changes but not in a homogeneous
way, so each site presents different displacement rates in envrionmental
spaces and as a consequence, different demographic rates. So we can
establish that for each site \(i\) in geographic space \(\mathbf{G}\),
there is a population size \(N_i\) that changes with respect to a
fitness function dependent on time-varying environmental conditions:

\[
N_i(\vec{e}_i(t)) \propto  S_i(\vec{e}_i(t))
\]

@@

Moreover, the effective population can be estimated as the harmonic mean
of the population in a demographic time series (Karlin 1968) for
\(\tau\) generations
\(\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ... + \frac{1}{N_\tau} } \Big)\)
for each \(i\) site in \(\mathbf{G}\). We substitute the harmonic mean
into the above equation, considering that although the number of
generations is not exactly equal to the time of the fixation index, they
are expected to be very close:

\[
F_{i}^{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{N_{ij}} 
\]

@@

We call this harmonic mean of environmental suitability
\emph{Suitability prevalence index }(\(S\)) given between \(0\) and
\(1\), since it reflects the rate of population growth as a function of
environment over a period of time. Thus we have a way to check that the
loss of heterozygosity (through the fixation index \(F_{st}\)) is
inversely proportional to the suitability prevalence for an \(i\) site
in the geography:

\[
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
\] If we consider all sites in the available geography, we call this
harmonic \emph{Suitability Prevalence Area}
\(\mathbf{S} = \{S_i | S_i > k\}\) since we expect that areas where the
\(F_{st}\) is lower are in areas where the prevalence over time is
higher, i.e.~environmental conditions remain suitable for a significant
period of time and therefore no effects of genetic drift of populations
have occurred. Conversely, it is expected that in populations with a
high fixation index value, fitness will not have prevailed over the same
time interval even though conditions currently exist and populations are
present.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Model the potential distribution of a taxon with present-day
  environmental conditions and obtain an environmental suitability map.
\item
  Project the potential distribution model to various environmental
  scenarios in the past with constant time steps and obtain the
  environmental suitability for each time scenario.
\item
  Characterize past environments and retain projections only for sites
  with analogous environmental conditions to the present.
\item
  Calculate the harmonic mean of the suitability for all the sites all
  time slices.
\end{enumerate}

@@

The average model obtained from Maxent was extrapolated to environmental
conditions into the past. Environmental layers were taken from pastclim
(Leonardi et al. 2023) in 2000 year time slices to cover a total of 120
000 years. For each scenario into the past, 10 replicates were
cross-validated and averaged.

@@

To calculate the historic endemic area, the SPA result is taken and a
threshold is applied to generate a binary map according to the
\(\mathbf{S}\) equation. This threshold is the proportion of time that
pixel has had ideal conditions. For our study, we consider the value of
90\% since for a period of 120 000 years it is a coverage of 100 000
years, which is a significantly evolutionary time to fix genetic
patterns at the population level.

@@

To test our hypothesis of genetic diversity as a function of
environmental suitability prevalence, from a set of genetically
structured populations, a matrix of fixation indices between populations
is considered. For this work we considerer the data from @Bono et al.
(2018). It consideres 10 populations along the known distribution and
provides values of \(F_{st}\) between this populations in matrix form.
The matrix was transformed into a table as pairs between populations
with index values between each pair. The values of the population with
itself, i.e.~\(0\), were included. Subsequently, georeferences were
obtained for each population considered in the fixation index matrix.
From this geo-referencing the SPA value for each population was
extracted. Finally, the SPA value was aligned in the table of fixation
index pairs. In total the table has 4 columns: source population, target
population, fixation index and SPA of the target population. This table
was separated by source population groups to generate a set of linear
models between SPA values and fixation indexes \(F_{st}\). An analysis
was performed for each source population with the hypothesis that the
correlation between SPA and fixation index exists and is non-zero if a
dispersal process has occurred from the historical endemic area of the
taxon.

@@

We show the species distribution model for \emph{Sciurus aberti} along
the time line of the projected area per time slice. We show 5 notable
moments in time line. and 5 notable points

@@

We get the suitability presence area (figure 2) delmited by a 0.5 and
0.9 threshold (Figure 4). The endemic region where the enviornmental
conditions 90 \% del tiempo is the Sierra Madre Occidental, in the
limits of Sonora, Chihuahua, Sinaloa and Durango. We correspond this
Area to the \emph{Sciurus aberti barbieri} subspecies. Also is notable
that at 0.5 threshold, other regions seems to have stable conditions. We
observe the area close to Coronado National Forest as well as Gila
National Forest and Mescalero Reservation as suitable areas along time
in the United States and some Areas of the Volcanic Belt in Mexico
(Puebla, north of Guerrero State and Jalisco.\\

@@

\hypertarget{fst-as-a-funciton-of-spa-and-fixation-index-projected-in-geography}{%
\subsubsection{Fst as a funciton of SPA and Fixation index projected in
geography}\label{fst-as-a-funciton-of-spa-and-fixation-index-projected-in-geography}}

@@

From the model we extrapolated the values from SPA along the geography
to new \(F_{st}\) values to get a geographical projection of fixation
index from \emph{S. a. barberi} population. (Figure 6)

@@

@@

>>>>

Our results reveal that fitness prevalence corresponds to the fixation
index of \emph{S. aberti} populations with respect to a source
population. Populations closer to the historical endemic area present a
higher genetic diversity and a lower \(F_{st}\) value. Finally, this
study allows us to add a biogeographic explanation to the results
obtained with population genetic methods and to generate maps of this
structure as tools to support conservation with a perspective that
integrates both population genetics and historical patterns of species
distribution.

@@

@@

We define \(\vec{e}_i =(e_1, e_2, ..., e_i )\) as a vector of
environmental variables, that is, a point in the space of \(v\)
environmental variables. Then \(R_i(\vec{e}_i)\) is a function that
relates fitness to environment, that is, for a species with
non-overlapping generations, the population's net reproductive rate
\(R_i\) is a measure of fitness that depends on environmental
conditions.

That is, the sites in the geography \(\mathbf{G}_o\) that present
positive growth rates are related to favorable conditions. If we take
the set of places in ambient space where these values are greater than a
threshold, we can define a volume in multidimensional space such that
\(R(\vec{e})\) is always positive and greater than this threshold, that
is \(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}\).

This is a way of expressing the fundamental niche as a function of
environmental conditions. So we can map each point of geography to
ambient space. Thus, for each site in the geography there will
correspond a site in the environmental space with different
environmental conditions and different growth rates. However all sites
within the region defined by \(\mathbf{N}_F\) will have viable
populations and positive rates. However, we assume that the environment
is dynamic and each geographic point is static, so for each site in
geographic space this growth rate depends on the environment and
therefore is different at each time.

Thus locally the growth rate will be related to the local population
size as long as it is within the boundary defined by \(\mathbf{N}_F\).
Therefore, the local population size as a function of the intrinsic rate
that depends on the change in environmental conditions over time is:
\(N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)\). That is, the locally effective
population size is a function of ideal environmental conditions.

We start by considering this defined volume of the set of \(R_0\) as a
set of trajectories for each point from geographic space to
environmental space. Considering this rate fixed \(\mathbf{R}_0\) when
the population grows, that is,
\(\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > 1 \}\) and therefore
any local population will have a positive growth rate.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,height=\textheight]{all_figures/figure_1.png}
\caption{Hypothetical niche in environmental space. An arbitrary point
moves within the ambient space entering and exiting the volume (niche)
where the intrinsic growth rate is positive.}
\end{figure}

@@

This set can be seen as a convex volume in environmental space and we
relate this growth rate to an environmental suitability function bounded
between \(0\) and \(1\), for example, for a unimodal multivariate pdf,
it would be equivalent to the distance from the maximum of the function.
That is, the fitness-related environmental suitability can be any
function such that it is \(0\) if \(R_i(\vec{e}_i) > 1\) and \(1\) when
\(R\) is evaluated at the expected value of \(\vec {e}\) for all values
of \(R_i(\vec{e}_i) > 1\). In a practical way, \(\mathbf{N}_F\) is
approximated from points of presence in the geography obtained from
databases, numerically approximating a convex surface with a surface
\(\mathbf{N}_R\) of the environmental space.

Once we have an approximation of this volume, we consider that each
point in the geography is static and has a trajectory in the
environmental space and that for each trajectory the fitness function
changes and therefore the suitability function changes but not
accordingly. in a homogeneous way, if not more so that each site
presents different displacement rates and different demographic rates.

The objective of this theoretical approach is not to know the shape of
the function of \(R_0\) but rather to find this local variation relating
this value of \(R_0\) and the function \(\mathbf{N}_F\) with a function
\$ S\_i(~vec\{e\}\_i)\$ of environmental suitability that can be
estimated from known methods of applied mathematics and machine learning
(for example the case of Maxent, Phillips, Anderson, and Schapire
(2006)).

Es decir:

\[
N_i \propto  S_i(\vec{e}_i)
\] Population size locally is a function of environmental suitability,
where if suitability is \(0\) the population is \(0\) and if suitability
is \(1\) the population has a maximum growth rate \(R_i(\vec{e}_i)\)

@@

Moreover, the effective population can be estimated as the harmonic mean
of the population in a demographic time series (Karlin 1968) for
\(\tau\) generations
\(\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ... + \frac{1}{N_\tau} } \Big)\).
We substitute the harmonic mean into the above equation, considering
that although the number of generations is not exactly equal to the time
of the fixation index, they are expected to be very close.

\[
F_{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{N_j} 
\]

@@

We call this harmonic mean of environmental suitability
\emph{Suitability prevalence index }(\(S\)) given between \(0\) and
\(1\), since it reflects the rate of population growth as a function of
environment over a period of time. Thus we have a way to check that the
loss of heterozygosity (through the fixation index \(F_{st}\)) is
inversely proportional to the fitness prevalence for an \(i\) site in
the geography:

\[
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
\] If we consider all sites in the available geography, we call this
harmonic mean of fitness \emph{Suitability prevalence Area}
\(\mathbf{S} = \{S_i | S_i > k\}\) since we expect that areas where the
\(F_{st}\) is lower are in areas where the prevalence over time is
higher, i.e.~environmental conditions remain suitable for a significant
period of time and therefore no effects of genetic drift of populations
have occurred. Conversely, it is expected that in populations with a
high fixation index value, fitness will not have prevailed over the same
time interval even though conditions currently exist and populations are
present.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  model the potential distribution of a taxon with present-day
  environmental conditions and obtain an environmental suitability map.
\item
  Project the potential distribution model to various environmental
  scenarios in the past with constant time steps and obtain the
  environmental suitability for each time scenario.
\item
  Characterize past analogous climates and retain projections only for
  sites with climates analogous to the present.
\item
  Generate a harmonic mean suitability map for all time slices.
\end{enumerate}

@@

The average model obtained from Maxent was extrapolated to environmental
conditions into the past. Environmental layers were taken from pastclim
(Leonardi et al. 2023) in 2000 year time slices to cover a total of 120
000 years. For each scenario into the past, 10 replicates were
cross-validated and averaged. For each projection into the past, as well
as the model of the present, we generated a binary map at a suitability
threshold of 90\% to get an estimate of the area distribution for each
time slice in pixels. In addition we fit a locally estimated scatterplot
smoothing (loess) model to the time series trend area using the R
package stats in order to observe patterns in the area - timeline
scatterplot.

@@

To calculate the historic endemic area, the SPA result is taken and a
threshold is applied to generate a binary map according to the
\(mathbf{S}\) equation. This threshold is the proportion of time that
pixel has had ideal conditions. For our study, we consider the value of
90\% since for a period of 120 000 years it is a coverage of 100 000
years, which is a significantly evolutionary time to fix genetic
patterns at the population level.

@@

To test our hypothesis of genetic diversity as a function of
environmental suitability prevalence, from a set of genetically
structured populations, a matrix of fixation indices between populations
is considered. For this work we considerer the data from @Bono et al.
(2018). It consideres 10 populations along the known distribution and
provides values of \(F_st\) between this populations in matrix form. The
matrix was transformed into a table as pairs between populations with
index values between each pair. The values of the population with
itself, i.e.~\(0\), were included. Subsequently, georeferences were
obtained for each population considered in the fixation index matrix.
From this geo-referencing the SPA value for each population was
extracted. Finally, the SPA value was aligned in the table of fixation
index pairs. In total the table has 4 columns: source population, target
population, fixation index and SPA of the target population. This table
was separated by source population groups to generate a set of linear
models between SPA values and fixation indexes \(F_{st}\). An analysis
was performed for each source population with the hypothesis that the
correlation between SPA and fixation index exists and is non-zero if a
dispersal process has occurred from the historical endemic area of the
taxon.

@@

We get the binary maps for all the models in suitability threshold of
0.9 and extract the number of pixels above this threshold in order to
get a relative predicted area. Here we show the trend from present
(Figure 3A) to past of the the projected area of \emph{Sciurus aberti}
along the time line (Figure 3F). The trend pattern exhibits an inverted
U-shape, showing that the present area is similar to the 120 000 area,
with a maximum in 22 000 coinciding with the Last Glacial Maximum in
North America (Figure 3B). We highlight three other notable scenarios,
the first local minimum after the interglacial maximum appears in the
year 30 000 (Figure 3C) coinciding with a distribution similar to the
present but with more suitable regions emerging in New Mexico and
Colorado; a local maximum in year 62 000 at the middle of the trend line
(Figure 3D), with a more suitable potential distribution towards the
south, with conditions disappearing practically everywhere in the United
States, except in the border with Sonora, but connecting the area of the
Sierra Madre Occidental with Zacatecas, San Luis Potosí and the
appearance of a corridor over the Transverse Volcanic Axis (from Jalisco
to Tlaxcala); finally we show a minimum area of the whole time series in
year 112 000. (Figure 3E), with the disappearance of conditions in
practically all the places where there were expansions, preserving
conditions only in some places of Chihuahua (Campo Verde and Tutuaca
Natural Protected Areas), Durango and Arizona (Prescott and Sedona).

@@

We get the suitability prevalence area delimited by a 0.5 and 0.9
threshold (Figure 4). The endemic region where the enviornmental
conditions 90 \% of the time is the Sierra Madre Occidental, in the
limits of Sonora, Chihuahua, Sinaloa and Durango. We correspond this
Area to the \emph{Sciurus aberti barbieri} subspecies. Also is notable
that at 0.5 threshold, other regions seems to have stable conditions. We
observe the area close to Coronado National Forest as well as Gila
National Forest and Mescalero Reservation as suitable areas along time
in the United States and some Areas of the Volcanic Belt in Mexico
(Puebla, north of Guerrero State and Jalisco.\\

@@

\hypertarget{f_st-as-a-funciton-of-spa-and-fixation-index-projected-in-geography}{%
\subsubsection{\texorpdfstring{\(F_{st}\) as a funciton of SPA and
Fixation index projected in
geography}{F\_\{st\} as a funciton of SPA and Fixation index projected in geography}}\label{f_st-as-a-funciton-of-spa-and-fixation-index-projected-in-geography}}

@@

From the model we extrapolated the values from SPA along the geography
to new \(F_{st}\) values to get a geographical projection of fixation
index from \emph{S. a. barberi} population. (Figure 6). The lowest
values are close to the Sierra Madre Occidental Area, in the limits of
Chihuahua and Sonora States as well as Sinaloa and Durango, score higher
near to the Tutuaca Wildlife regue There is also a hotspot between
Sonora And arizona, near to Bisbee Arizona in United States as well as
Naco Sonora y Mexico.

@@

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

With the current availability of occurrence data and models of potential
distribution projected to past climate scenarios it is possible to find
geographic patterns of differentiation in populations at an acceptable
resolution (half a degree) that allows us to define sites where the
species can potentially be found but share different degrees of genetic
diversity.

Projections of genetic diversity in geography have been made in previous
studies Van Zonneveld et al. (2012). However in a general way they are
extrapolations of observations, without a predictive model, which with
some explanatory variable behind can explain in a statistically
convincing way the information obtained from population genetics.

We have shown a generalizable protocol for predicting patterns of
population genetics in geography. In this way we link information on the
evolutionary history of a taxon and its ecology. In addition, we provide
a predictive vehicle that generates new geographic information
applicable to decision making in both survey conduct and taxon
conservation.

Thanks to the theoretical support that we have briefly outlined in this
paper, it is possible to explain the explanatory value of the
Coefficient of determination for our linear model (\$R\^{}2 = 0.79 \$),
which is information coming from two completely different sources,
without any correlation, but always guided by the premise of the
evolutionary process as causal.

These maps represent a turning point in the conception of conservation.
With the tool we propose in this study, it is possible to delimit
conservation sites based on evidence of current and historical
distribution and genetic diversity among different populations.

While many conservation decisions are made on the basis of genetic
diversity, studying the genetic structure of populations at a
macro-scale is often very complicated, and making conservation decisions
based on current distribution without taking into account the historical
distribution of taxa can lead to a misguided conservation strategy.

The generalization of our protocol and its implementation in tools
widely used by biogeography allow its direct implementation in many case
studies with available information, as well as the design of new
questions, e.g.~species assemblages under prevalence of environmental
suitability. A research path was also established on whether the process
of environmental stability over time is related to biodiversity.

@@

\leavevmode\vadjust pre{\hypertarget{ref-van2012mapping}{}}%
Van Zonneveld, Maarten, Xavier Scheldeman, Pilar Escribano, Maria A
Viruel, Patrick Van Damme, Willman Garcia, César Tapia, José Romero,
Manuel Siguenas, and José I Hormaza. 2012. {``Mapping Genetic Diversity
of Cherimoya (Annona Cherimola Mill.): Application of Spatial Analysis
for Conservation and Use of Plant Genetic Resources.''} \emph{PloS One}
7 (1): e29845.
